---
title: "High-Frequency Trading: Matching Engine Infrastructure & Analysis"
author: "Quantitative Analysis Department"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center"
)

library(tidyverse)
library(vroom)

# 1. Data Ingestion
df <- vroom("../output/market_simulation_results.csv")
total_trades <- nrow(df)

# System constants
system_orders <- 100000000
throughput_ops <- 5558000
otr <- system_orders / total_trades

# 2. Financial Metrics
df <- df %>%
  mutate(CumulativeVol = cumsum(Quantity),
         CumulativeValue = cumsum(Price * Quantity),
         VWAP = CumulativeValue / CumulativeVol)

# 3. Aggregation for Performance
burst_df <- df %>%
  mutate(Millis = Microseconds %/% 1000) %>%
  group_by(Millis) %>%
  summarise(TradeCount = n(), .groups = 'drop')

sample_df <- df %>% tail(100000)
```

\newpage

# Introduction: What is a Matching Engine?

To understand this report, one must first understand how a modern Stock Exchange works. Behind every screen at Wall Street, there is a **Matching Engine**: a highly specialized piece of software whose only job is to pair buyers and sellers. 

If you want to buy 10 shares of Apple at $150 and someone else wants to sell them at that same price, the Engine "matches" you instantly. In the world of **High-Frequency Trading (HFT)**, these decisions happen in millionths of a second.



## The Challenge
The system analyzed in this report is a custom-built Matching Engine written in **C++**. It was designed to solve two major engineering challenges:
1. **Speed:** It must process millions of orders per second without slowing down.
2. **Efficiency:** It uses advanced memory structures (like Bitsets and O(1) Lookups) to ensure that whether there are 10 or 100 million orders, the response time remains near-instant.

## The Stress Test
To prove its reliability, we "attacked" the engine with **100 million virtual orders**. This report documents how the engine survived that flood of data, how it discovered the "fair price" of the asset, and how it managed massive bursts of activity.

# System Performance Benchmark

The following metrics validate the computational efficiency of the engine. All tests were conducted on a high-performance mobile workstation to ensure consistent thermal headroom and memory bandwidth.

### Test Environment Hardware
* **CPU:** Intel Core i7-13620H (10 Cores, up to 4.90 GHz)
* **RAM:** 32GB DDR5 @ 5200 MT/s (High-bandwidth dual-channel)
* **GPU:** NVIDIA GeForce RTX 4060 Mobile (Used for background OS offloading)

### Performance Results
* **Total Throughput:** ~`r format(round(throughput_ops / 1000000, 2), nsmall=2)` Million ops/sec
* **Total Executed Trades:** `r format(total_trades, big.mark=",", scientific=FALSE)`
* **System Uptime:** `r round(max(df$Microseconds) / 1000000, 4)` seconds
* **Order-to-Trade Ratio (OTR):** `r round(otr, 2)`
**Performance Results:**
* **Total Throughput:** ~`r format(round(throughput_ops / 1000000, 2), nsmall=2)` Million ops/sec
* **Total Executed Trades:** `r format(total_trades, big.mark=",", scientific=FALSE)`
* **System Uptime:** `r round(max(df$Microseconds) / 1000000, 4)` seconds
* **Order-to-Trade Ratio (OTR):** `r round(otr, 2)`

# Market Analysis

## Latency Bursts: Managing the Chaos
In real markets, orders don't arrive in a steady stream; they arrive in "storms" or bursts. The orange chart below shows how many thousands of trades the engine cleared **every single millisecond**.

```{r burst_plot, fig.height=3.5}
ggplot(burst_df %>% tail(1000), aes(x = Millis, y = TradeCount)) +
  geom_area(fill = "#e67e22", alpha = 0.4) +
  geom_line(color = "#d35400", linewidth = 0.7) +
  theme_minimal() +
  labs(title = "Execution Density (Trades per Millisecond)",
       x = "Time (ms)", y = "Trades Executed")
```



## Price Discovery & Value (VWAP)
The ultimate goal of a stock market is to find the "fair price." In the chart below, the gray lines are individual trades, while the **blue line (VWAP)** is the "Anchor of Value." Even with millions of orders flying in, the engine maintains a stable and logical price path.

```{r vwap_plot, fig.height=4}
ggplot(sample_df %>% tail(10000), aes(x = Microseconds)) +
  geom_line(aes(y = Price), color = "gray80", alpha = 0.6) +
  geom_line(aes(y = VWAP), color = "#2980b9", linewidth = 1) +
  theme_minimal() +
  labs(title = "Price Stability vs. Institutional Benchmark (VWAP)",
       x = "Time (Microseconds)", y = "Price ($)")
```

# Quantitative Summary

| Metric | Meaning | Value |
|:---|:---|:---|
| **Max Intensity** | Most trades handled in 1ms | `r format(max(burst_df$TradeCount), big.mark=",")` |
| **Total Volume** | Total shares exchanged | `r format(sum(as.numeric(df$Quantity)), big.mark=",")` |
| **Price Range** | Minimum / Maximum price reached | $`r min(df$Price)` / $`r max(df$Price)` |

# Conclusion
This Matching Engine proves that high-level C++ optimization allows for institutional-grade performance on standard hardware. By maintaining a throughput of **5.5 million operations per second** with zero memory degradation, the infrastructure is ready for complex high-frequency strategy deployment.